---
geometry: margin = 1in
output: pdf_document
---

\pagenumbering{gobble}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
yt <- 0
```

# Working with continuour prior distributions
## Stat 340, Fall 2021

\bigskip

### Your turn `r (yt <- yt+1)`: Undertanding kernels

The kernel of the Beta distribution is $x^{a-1}(1 - x)^{b-1}$. To help you develop your understanding of the kernel and how it relates to the PDF, we will compare plots of the beta kernel to the full PDF.

1. Pick a value for the `a` and `b` parameters.

\vspace{1.25in}

2. Using these parameters, run the below code chunk. (This code chunk is posted on teh course webpage, so just copy and paste it to save time!)

    ```{r eval=FALSE}
a <- # insert your a value
b <- # insert your b value
x <- seq(0, 1, by = .001) 
kernel <- x^{a - 1} * (1 - x)^{b - 1} 
pdf <- gamma(a + b)/(gamma(a) * gamma(b)) * kernel 
par(mfrow=c(1,2))
plot(x, pdf, type="l", main = "PDF")
plot(x, kernel, type = "l", main = "Kernel")
```

3. What are the similarities between the pdf and the kernel? What are the differences?

\vspace{1.25in}

4. If we know the kernel for a distribution, why can we find the PDF?

\vspace{1.25in}

\clearpage

### Your turn `r (yt <- yt+1)`: Geometric-Beta model

Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from the geometric distribution with PMF 
$$f(x| p) = p(1-p)^{x-1};\ x = 1, 2, \ldots, \ \ 0 < p < 1$$
1. Write down the likelihood function, $f(x_1, \ldots, x_n | \lambda)$. (Remember joint distributions!)

\vspace{1.25in}

2. Suppose that you decide to use a Beta(a,b) prior distribution for $p$ with PDF

    $$\pi(p) = \frac{\Gamma(a + b)}{\Gamma(a) \Gamma(b)} p^{a-1} (1-p)^{b-1}$$ 

    Find the posterior density of $p$.
    
    \vspace{3.75in}

3. Is the beta prior a conjugate family to the geometric likelihood?


