---
title: "MCMC software"
author: "Stat 340: Bayesian Statistics"
output:
  xaringan::moon_reader:
    css: ["default", "assets/css/my-theme.css", "assets/css/my-fonts.css", "hygge"]
    seal: false
    lib_dir: libs
    nature:
      output:
      ratio: '16:9'
      highlightStyle: solarized-light
      highlightLanguage: ["r", "css", "yaml"]
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
xaringanExtra::use_tachyons()
library(runjags)
library(coda)
library(countdown)

```

class: middle, clear

class: title-slide, left, middle


# `r rmarkdown::metadata$title`

### `r rmarkdown::metadata$author`

---

class: inverse

# Your turn



---

# Summary of MCMC

- With the combination of Gibbs and Metropolis
sampling we can fit virtually any model

- In some cases Bayesian computing is actually preferable
to maximum likelihood analysis

- In most cases Bayesian computing is slower

- However, it is often worth the wait for improved uncertainty quantification and interpretability

- In all cases it is important to carefully monitor convergence 

.footnote[Slide adapted from Brian Reich]

---

# Why Just Another Gibbs Sampler (JAGS)?

- You can fit virtually any model

- You can call JAGS from R, allowing for plotting and
data manipulation in R

- It runs on all platforms: LINUX, Mac, Windows

- There is a lot of help online

- R has many built in packages for convergence diagnostics

.footnote[Slide adapted from Brian Reich]

---

# How does JAGS work?


- You specify the model by declaring the likelihood and priors

- JAGS then sets up the MCMC sampler
    + e.g., works out the full conditional distributions for all parameters

- It returns MCMC samples in a matrix or array

- It also automatically produces posterior summaries like
means, credible sets, and convergence diagnostics

- User’s manual: http://people.stat.sc.edu/hansont/stat740/jags_user_manual.pdf


.footnote[Slide adapted from Brian Reich]

---

# Steps for any analysis using JAGS


1. Specify the model as a string (or script)

2. Define the data and the prior parameters

3. Define initial values

4. Draw posterior samples using `run.jags()` 

5. Inspect/summarize the results using the `plot()` and `summary()`


---

# Example: Rocket launches

$Y_i =$ 0 or 1 (failure/success)

.bold[Likelihood:] $Y_i \overset{\rm iid}{\sim}{\rm Bernoulli} (\theta)$

.bold[Prior:] Elicitation leads to uniform on (0.1, 0.9)

.bold[Posterior:]

$$p(\theta | y) \propto \begin{cases}
\theta^3 (1-\theta)^8 & \text{if } 0.1 < \theta < 0.9\\
0 & \text{otherwise.}\end{cases}$$


---

## 0. Load runjags

```{r message=FALSE}
library(runjags)
```

---

## 1. Specify the model as a string

```{r}
model_string <- "
model{

  # Likelihood
  for(i in 1:n) {
    y[i] ~ dbern(theta)
  }
  
  # Prior
  theta ~ dunif(a, b)
}
"
```


---

## 2. Load the data and define prior parameters

Loading the launch data
```{r}
launches <- read.table(
  "https://alysongwilson.github.io/BR/table21.txt", 
  header = TRUE
)
```

Define list with data and prior parameters
```{r}
launch_data <- list(
  y = launches$Outcome,
  n = nrow(launches),
  a = 0.1,
  b = 0.9
)
```

---

## 3. Define initial values


---

## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
]

.right-narrow[
```{r eval=FALSE}
posterior <- run.jags(
  model = model_string, #<<
  n.chains = 1,
  data = launch_data,
  monitor = "theta",
  adapt = 1000,
  burnin = 5000,
  sample = 5000
)
```
]

---

## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
- `n.chains` the number of chains to run
]

.right-narrow[
```{r eval=FALSE}
posterior <- run.jags(
  model = model_string, 
  n.chains = 1, #<<
  data = launch_data,
  monitor = "theta",
  adapt = 1000,
  burnin = 5000,
  sample = 5000
)
```
]

---

## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
- `n.chains` the number of chains to run
- data a named list or data frame include the data and prior parameter values
]

.right-narrow[
```{r eval=FALSE}
posterior <- run.jags(
  model = model_string, 
  n.chains = 1, 
  data = launch_data, #<<
  monitor = "theta",
  adapt = 1000,
  burnin = 5000,
  sample = 5000
)
```
]

---

## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
- `n.chains` the number of chains to run
- `data` data a named list or data frame include the data and prior parameter values
- `monitor` character vector of the names of variables to monitor
]

.right-narrow[
```{r eval=FALSE}
posterior <- run.jags(
  model = model_string, 
  n.chains = 1, 
  data = launch_data, 
  monitor = "theta", #<<
  adapt = 1000,
  burnin = 5000,
  sample = 5000
)
```
]

---


## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
- `n.chains` the number of chains to run
- `data` data a named list or data frame include the data and prior parameter values
- `monitor` character vector of the names of variables to monitor
- `adapt` number of samples drawn during initial sampling/adaptation phase
]

.right-narrow[
```{r eval=FALSE}
posterior <- run.jags(
  model = model_string, 
  n.chains = 1, 
  data = launch_data, 
  monitor = "theta",
  adapt = 1000,  #<<
  burnin = 5000,
  sample = 5000
)
```
]

---


## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
- `n.chains` the number of chains to run
- `data` data a named list or data frame include the data and prior parameter values
- `monitor` character vector of the names of variables to monitor
- `adapt` number of samples drawn during initial sampling/adaptation phase
- `burnin` number of burnin iterations, NOT including the adaptive iterations
]

.right-narrow[
```{r eval=FALSE}
posterior <- run.jags(
  model = model_string, 
  n.chains = 1, 
  data = launch_data, 
  monitor = "theta",
  adapt = 1000,  
  burnin = 5000, #<<
  sample = 5000
)
```
]

---



## 4. Draw posterior samples using `run.jags()` 

.left-wide[
- `model` string specifying the model
- `n.chains` the number of chains to run
- `data` data a named list or data frame include the data and prior parameter values
- `monitor` character vector of the names of variables to monitor
- `adapt` number of samples drawn during initial sampling/adaptation phase
- `burnin` number of burn-in iterations, NOT including the adaptive iterations
- `sample` the total number of (additional) samples to draw
]

.right-narrow[
```{r run-jags-launches, eval=FALSE}
posterior <- run.jags(
  model = model_string, 
  n.chains = 1, 
  data = launch_data, 
  monitor = "theta",
  adapt = 1000,  
  burnin = 5000, 
  sample = 5000 #<<
)
```
]

---
class: clear

In an .Rmd file, add `silent.jags = TRUE` to avoid this issue...

```{r run-jags-launches, eval=TRUE, echo=FALSE}
```



---

```{r plot-posterior, fig.height = 4.5, fig.width = 8, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
svg("figs/plot-launch-posterior.svg", height = 4.5, width = 8)
plot(posterior)
dev.off()
```

background-image: url("figs/plot-launch-posterior.svg")
background-size: 85%

## 5. `plot(posterior)`



---


class: clear

`runjags` objects are big lists

```{r}
names(posterior)
```

---

`mcmc` element is a list of the actual draws

```{r}
head(posterior$mcmc)
```

---

## 5. Inspect/summarize the results

```{r}
summary(posterior$mcmc[[1]])
```


---

## `bayesplot`

- provides a variety of `ggplot2`-based plotting functions for use after fitting Bayesian models

- Visualizations of MCMC simulations and diagnostics *from any algorithm* (`mcmc_*`)

- Graphical prior and posterior predictive checks (`ppc_*`)

---

## Convert to an `mcmc` object

`bayesplot` requires objects to be of class `mcmc`

You'll need to convert your output from `run.jags()`

```{r collapse=TRUE}
class(posterior)
```

```{r collapse=TRUE}
post_mcmc <- as.mcmc(posterior)
class(post_mcmc)
```

---

## Trace plot

```{r fig.height = 3.5, fig.width = 5, out.width = "57%", fig.align='center'}
mcmc_trace(post_mcmc)
```

---

## ACF plot

```{r fig.height = 3.5, fig.width = 5, out.width = "57%", fig.align='center'}
mcmc_acf(post_mcmc)
```

---

## Posterior density plot

```{r fig.height = 3.5, fig.width = 5, out.width = "57%", fig.align='center'}
mcmc_dens(post_mcmc)
```

---

## Plot equal-tail interval estimates

```{r fig.height = 3.5, fig.width = 5, out.width = "57%", fig.align='center'}
mcmc_areas(post_mcmc)
```

---

class: inverse

# Your turn

.Large[.bold[How do we calculate the posterior probability that the proportion of successful launches is at least 0.5?]]

.Large[.bold[Discuss with your neighbor.]]


```{r echo=FALSE}
countdown(1, 30)
```


---

## Posterior inference for `runjags` objects

What's the posterior probability that the proportion of successful launches is at least 0.5?

```{r collapse=TRUE}
mean(posterior$mcmc[[1]] >= 0.5)
```

--

95% credible interval for $\theta$, the proportion of successful launches

```{r collapse=TRUE}
quantile(posterior$mcmc[[1]], probs = c(0.05, 0.95))
```





---

class: inverse

# Your turn

.Large[.bolder[Work through the examples on the JAGS handout with a neighbor.]]

.Large[.bolder[Check GitHub for a .Rmd template]]

.Large[.bolder[https://github.com/aloy/math315-fall2019]]

