---
title: "Bayesian linear regression"
subtitle: "Model fitting in JAGS"
author: "Stat 340: Bayesian Statistics"
output:
  xaringan::moon_reader:
    css: ["default", "assets/css/my-theme.css", "assets/css/my-fonts.css", "hygge"]
    seal: false
    lib_dir: libs
    nature:
      output:
      ratio: '16:9'
      highlightStyle: solarized-light
      highlightLanguage: ["r", "css", "yaml"]
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
library(gridExtra)
library(ggplot2)
library(ggthemes)
library(LearnBayes)
library(bayesrules)
library(runjags)
library(coda)
library(bayesplot)
library(patchwork)
library(tidyverse)
library(ggridges)
library(countdown)
xaringanExtra::use_tachyons()

Howell1 <- readr::read_delim("https://raw.githubusercontent.com/rmcelreath/rethinking/master/data/Howell1.csv", delim = ";")
adults <- filter(Howell1, age >= 18)

```

class: title-slide, left, middle

# `r rmarkdown::metadata$title`:<br> `r rmarkdown::metadata$subtitle`

### `r rmarkdown::metadata$author`


---

## JAGS for Bayesian SLR

Write down the model string

```{r}
slr_model <-"model {
## sampling model
for (i in 1:N){
y[i] ~ dnorm(beta0 + beta1 * x[i], invsigma2)
}

## priors
beta0 ~ dnorm(mu0, g0)
beta1 ~ dnorm(mu1, g1)
invsigma2 ~ dgamma(a, b)
sigma <- sqrt(pow(invsigma2, -1))
}"
```

---

## JAGS for Bayesian SLR

Define the data and set prior parameters

```{r}
the_data <- list(
  y = adults$height,   # response variable
  x = adults$weight,   # explanatory variable
  N = nrow(adults),    # sample size
  mu0 = 0,             # prior mean for beta0
  g0 = 0.0001,         # prior precision for beta0
  mu1 = 0,             # prior mean for beta1
  g1 = 0.0001,         # prior precision for beta1
  a = 1,               # prior shape for 1/sigma2
  b = 1                # prior scale for 1/sigma2
)
```

---

## JAGS for Bayesian SLR

Generate samples from the posterior

```{r warning=FALSE, message=FALSE, results='hide'}
posterior <- run.jags(
  slr_model,
  data = the_data,
  n.chains = 1,
  monitor = c("beta0", "beta1", "sigma"),
  adapt = 1000,
  burnin = 5000,
  sample = 5000,
  silent.jags = TRUE
)
```

---

## Summary of the fitted model

```{r}
summary(posterior)
```

---

## MCMC diagnostics

```{r fig.height = 3, fig.width = 9, out.width = "100%"}
mcmc_trace(posterior$mcmc)
```

---

## MCMC diagnostics

```{r fig.height = 3, fig.width = 9, out.width = "100%"}
mcmc_acf(posterior$mcmc)
```

---

## Setting `thin = 50`

```{r warning=FALSE, message=FALSE, results='hide'}
posterior <- run.jags(
  slr_model,
  data = the_data,
  n.chains = 1,
  monitor = c("beta0", "beta1", "sigma"),
  adapt = 1000,
  burnin = 5000,
  sample = 5000,
  thin = 50,
  silent.jags = TRUE
)
```

---

## MCMC diagnostics

```{r fig.height = 3, fig.width = 9, out.width = "100%"}
mcmc_trace(posterior$mcmc)
```

---

## MCMC diagnostics

```{r fig.height = 3, fig.width = 9, out.width = "100%"}
mcmc_acf(posterior$mcmc)
```

---

## Summary of the fitted model

```{r}
summary(posterior)
```

---

## Plotting the fitted model

.pull-left[
```{r}
post_means <- apply(
  posterior$mcmc[[1]], 2, mean
)
```
]

.pull-right[
```{r echo=FALSE, fig.height=4, fig.width=4, out.width = "90%"}
ggplot(data = adults) +
  geom_point(mapping = aes(x = weight, y = height), alpha = 0.5) +
  geom_abline(intercept = post_means["beta0"], slope = post_means["beta1"], color = "steelblue", size = 1) +
  theme_light()
```

]


---

## Sampling from the joint posterior

.pull-left[
```{r}
post_draws <- as.data.frame(
  posterior$mcmc[[1]]
)
head(post_draws)
```
]

.pull-right[
```{r echo=FALSE, fig.height=4, fig.width=4, out.width = "90%"}
ggplot(data = adults) +
  geom_point(mapping = aes(x = weight, y = height), alpha = 0.5) +
  geom_abline(data = post_draws[1:1000,], mapping = aes(intercept = beta0, slope = beta1), alpha = 0.1, color = "steelblue") +
  labs(title = "1000 draws from joint posterior") +
  theme_light()
```

]

---

## Generating mean responses

.pull-left[
.code100[
```{r}
mu_at_50 <- with(post_draws, 
                 beta0 + beta1 * 50)
```
]

```{r fig.height=3, fig.width=4, echo=FALSE, out.width="65%", fig.align='center'}
mcmc_areas(data.frame(mu50 = mu_at_50), prob = .9)
```

.code100[
```{r collapse=TRUE}
quantile(mu_at_50, probs = c(0.05, 0.95))
```
]

]

.pull-right[
```{r fig.height=4, fig.width=4, echo=FALSE, out.width="90%"}
ggplot(data = adults) +
  geom_point(mapping = aes(x = weight, y = height), alpha = 0.5) +
  geom_abline(intercept = post_means["beta0"], slope = post_means["beta1"], color = "blue") +
  geom_errorbar(mapping = aes(x = 50, ymin = quantile(mu_at_50, probs = 0.05), ymax = quantile(mu_at_50, probs = 0.95)), color = "orange") +
  labs(title = "1000 draws from joint posterior") +
  theme_light()
```
]

---

## Generating predicted responses

.pull-left[
.code100[
```{r}
mu_at_50 <- with(post_draws, 
                 beta0 + beta1 * 50)
S <- length(mu_at_50)
y_at_50 <- rnorm(S, 
                 mu_at_50, 
                 post_draws$sigma)
```
]

```{r fig.height=3, fig.width=4, echo=FALSE, out.width="65%", fig.align='center'}
mcmc_areas(data.frame(y50 = y_at_50), prob = .9)
```

]

.pull-right[
```{r fig.height=4, fig.width=4, echo=FALSE, out.width="90%"}
ggplot(data = adults) +
  geom_point(mapping = aes(x = weight, y = height), alpha = 0.5) +
  geom_abline(intercept = post_means["beta0"], slope = post_means["beta1"], color = "blue") +
  geom_errorbar(mapping = aes(x = 50, ymin = quantile(y_at_50, probs = 0.05), ymax = quantile(y_at_50, probs = 0.95)), color = "orange") +
  labs(title = "1000 draws from joint posterior") +
  theme_light()
```
]


---
class: inverse

# Your turn

.bold[
Work through the Bayesian regression handout with your neighbor(s). 

Think about what ideas in simple linear regression are still unclear.
]